Longitudinal Language Analysis in the Diagnosis of Mild Cognitive Impairment: A Case Study

### 2019-02-07 Source/Presidents Corpus Paper/paper1.tex
Paper 1 - Rough Draft

% HEADER BEGINS
\documentclass[10pt]{article}
%% Packages
\input{settings/packages}
\usepackage{graphicx}
\usepackage{float}
\usepackage{enumitem}
\usepackage{fancyhdr}
\usepackage{rotating}
%% Page Settings
\input{settings/page}
%% Own Commands
\pagestyle{fancy}
\fancyhf{}
\fancyhead[CE,CO]{Early Detection of MCI that progresses to dementia - An Interdisciplinary Approach}
\fancyfoot[LE,RO]{\thepage}

\renewcommand{\headrulewidth}{2pt}
\renewcommand{\footrulewidth}{1pt}
\input{settings/macros}
\title{Qualifying Report}
\date{2019-01-31}
\author{Jomar Alcantara}
% HEADER ENDS
% DOCUMENT BEGINS
\begin{document}
\section{Abstract}
\section{Background}
Dementia has been identified as one of the fastest growing difficulties facing the world today. A recent report suggests that in 2015 there were 46 million people with a diagnosis of dementia and that number is expected to hit 131.5 million by 2050 \cite{Prince2015}. The report also states that the worldwide cost of dementia in 2018 is estimated to be in the region of one trillion US dollars.
\par
In 2009, the Department of Health designed it's National Dementia strategy and as part of this made early diagnosis and support one of it's key priorities \cite{England2009}. A lot of work has gone into trying to find ways of improving the early diagnosis of Alzheimer's Disease (AD) and Mild Cognitive Impairment (MCI) with research focused on two areas - identifying biological markers and analyzing the cognitive decline of those who are suspected to have the disease \cite{Taler2008}. As described above, the numbers of those suffering from AD and MCI are going to increase as the population ages \cite{Prince2015} and thus it is important that we utilize technology wherever possible to aid clinicians in the detection of MCI and AD. At the present time diagnosis is typically conducted at memory clinics by trained clinicians \cite{Boschi2017}. I theorize that we may be able to enable an earlier diagnosis of those with MCI and AD using samples of spontaneous speech, natural language processing (NLP) and machine learning (ML).
\par
There is a large body of research that looks at the decline in language of those with MCI and AD \cite{Taler2008, Boschi2017}. However there is conflicting evidence in these studies about which declining language factors are associated with MCI and AD \cite{Taler2008, Boschi2017}. Research therefore should look at these features in more detail and a clarification of this currently disorganised picture should go some way to helping researchers further understand the disease and it's progression. There is some evidence to support this, Berisha et al \cite{Berisha2015}, has shown through a longitudinal language analysis of spontaneous speech that there are marked differences in different parts of speech in this process between those who would go on to have a diagnosis of AD and a healthy control. 
\par
The potential impact of this research in this area is immense. Research has shown that early diagnosis of people with AD or MCI improves sufferers quality of life and can, in some cases, slow the progress of the disease. Early diagnosis can increase the number of research opportunities for understanding the early stages of dementia and how the disease progresses so that more research can be conducted which may, in the future, lead to new treatments and other interventions.

@@

\bibliographystyle{unsrt}
\bibliography{QualifyingReport}

>>>>

% HEADER BEGINS
\documentclass[10pt]{article}
%% Packages
\input{settings/packages}
\usepackage{graphicx}
\usepackage{float}
\usepackage{enumitem}
\usepackage{fancyhdr}
\usepackage{rotating}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}

%% Page Settings
\input{settings/page}
%% Own Commands
\input{settings/macros}
\title{Longitudinal Language Analysis in the Diagnosis of Mild Cognitive Impairment: A Case Study}

\author{
  Jomar C. Alcantara \\
  Department of Computer Science \\
  School of Engineering and Applied Sciences\\
  Aston University\\
  Birmingham, UK \\
  \texttt{alcantaj@aston.ac.uk} \\
  \And
  Peter Sawyer \\
  Department of Computer Science \\
  School of Engineering and Applied Sciences \\
  Aston University\\
  Birmingham, UK \\
  \texttt{p.sawyer@aston.ac.uk} \\
  \And
  Cristina Romani \\
  Department of Psychology \\
  School of Life and Health Science\\
  Aston University\\
  Birmingham, UK \\
  \texttt{c.romani@aston.ac.uk} \\
}
% HEADER ENDS
% DOCUMENT BEGINS
\begin{document}
\maketitle
\section{Abstract}
\section{Background}
Dementia has been identified as one of the fastest growing difficulties facing the world today. A recent report suggests that in 2015 there were 46 million people with a diagnosis of dementia and that number is expected to hit 131.5 million by 2050 \cite{Prince2015}. The report also states that the worldwide cost of dementia in 2018 is estimated to be in the region of one trillion US dollars.
In 2009, the Department of Health designed it's National Dementia strategy and as part of this made early diagnosis and support one of it's key priorities \cite{England2009}. A lot of work has gone into trying to find ways of improving the early diagnosis of Alzheimer's Disease (AD) and Mild Cognitive Impairment (MCI) with research focused on two areas - identifying biological markers and analyzing the cognitive decline of those who are suspected to have the disease \cite{Taler2008}. As described above, the numbers of those suffering from AD and MCI are going to increase as the population ages \cite{Prince2015} and thus it is important that we utilize technology wherever possible to aid clinicians in the detection of MCI and AD. At the present time diagnosis is typically conducted at memory clinics by trained clinicians \cite{Boschi2017}. I theorize that we may be able to enable an earlier diagnosis of those with MCI and AD using samples of spontaneous speech, natural language processing (NLP) and machine learning (ML).
There is a large body of research that looks at the decline in language of those with MCI and AD \cite{Taler2008, Boschi2017}. However there is conflicting evidence in these studies about which declining language factors are associated with MCI and AD \cite{Taler2008, Boschi2017}. Research therefore should look at these features in more detail and a clarification of this currently disorganised picture should go some way to helping researchers further understand the disease and it's progression. There is some evidence to support this, Berisha et al \cite{Berisha2015}, has shown through a longitudinal language analysis of spontaneous speech that there are marked differences in different parts of speech in this process between those who would go on to have a diagnosis of AD and a healthy control. 
The potential impact of this research in this area is immense. Research has shown that early diagnosis of people with AD or MCI improves sufferers quality of life and can, in some cases, slow the progress of the disease. Early diagnosis can increase the number of research opportunities for understanding the early stages of dementia and how the disease progresses so that more research can be conducted which may, in the future, lead to new treatments and other interventions.

@@

\begin{thebibliography}{10}

\bibitem{Prince2015}
Martin Prince, Anders Wimo, Guerchet M, Ali GC, Wu~YT, and Prina M.
\newblock {World Alzheimer Report 2015 The Global Impact of Dementia An
  analysis of prevalence, incidence, cost and trends}.
\newblock {\em Alzheimer's Disease International}, 2015.

\bibitem{England2009}
{Department of Health}.
\newblock {Living well with dementia : Dementia Strategy (England)}.
\newblock Technical report, Department of Health, 2009.

\bibitem{Taler2008}
Vanessa Taler and Natalie~A. Phillips.
\newblock {Language performance in Alzheimer's disease and mild cognitive
  impairment: A comparative review}.
\newblock {\em Journal of Clinical and Experimental Neuropsychology},
  30(5):501--556, 2008.

\bibitem{Boschi2017}
Veronica Boschi, Eleonora Catrical{\`{a}}, Monica Consonni, Cristiano Chesi,
  Andrea Moro, and Stefano~F. Cappa.
\newblock {Connected speech in neurodegenerative language disorders: A review}.
\newblock {\em Frontiers in Psychology}, 8(MAR), 2017.

\bibitem{Berisha2015}
Visar Berisha, Shuai Wang, Amy LaCross, and Julie Liss.
\newblock {Tracking Discourse Complexity Preceding Alzheimer's Disease
  Diagnosis: A Case Study Comparing the Press Conferences of Presidents Ronald
  Reagan and George Herbert Walker Bush}.
\newblock {\em Journal of Alzheimer's Disease}, 45(3):959--963, 2015.

\bibitem{Orimaye2014}
Sylvester~Olubolu Orimaye, Jojo Sze-Meng Wong, and Karen~Jennifer Golden.
\newblock {Learning Predictive Linguistic Features for Alzheimer's Disease and
  related Dementias using Verbal Utterances}.
\newblock {\em Proceedings of the Workshop on Computational Linguistics and
  Clinical Psychology: From Linguistic Signal to Clinical Reality}, pages
  78--87, 2014.

\bibitem{Fraser2015}
Kathleen~C. Fraser, Jed~A. Meltzer, and Frank Rudzicz.
\newblock {Linguistic Features Identify Alzheimer's Disease in Narrative
  Speech}.
\newblock {\em Journal of Alzheimer's Disease}, 49(2):407--422, oct 2015.

\bibitem{Bird2009}
Steven Bird, Ewan Klein, and Edward Loper.
\newblock {\em {Language Processing and Python}}.
\newblock O'Reilly Media Inc, 2009.

\bibitem{Richards1987}
Brian Richards.
\newblock {Type/Token Ratios: what do they really tell us?}
\newblock {\em Journal of Child Language}, 14(02):201, jun 1987.

\bibitem{Bird2000}
Helen Bird, Matthew~A. {Lambon Ralph}, Karalyn Patterson, and John~R. Hodges.
\newblock {The rise and fall of frequency and imageability: Noun and verb
  production in semantic dementia}.
\newblock {\em Brain and Language}, 73(1):17--49, 2000.

\bibitem{Le2011}
Xuan Le, Ian Lancashire, Graeme Hirst, and Regina Jokel.
\newblock {Longitudinal detection of dementia through lexical and syntactic
  changes in writing: A case study of three British novelists}.
\newblock {\em Literary and Linguistic Computing}, 26(4):435--461, 2011.

\bibitem{Garrard2005}
Peter Garrard, Lisa~M. Maloney, John~R. Hodges, and Karalyn Patterson.
\newblock {The effects of very early Alzheimer's disease on the characteristics
  of writing by a renowned author}.
\newblock {\em Brain}, 2005.

\bibitem{Almor1999}
Amit Almor, Daniel Kempler, Maryellen~C MacDonald, Elaine~S Andersen,
  Lorraine~K Tyler, Lee Willis, Lori Altmann, Mariela Gil, Laila Lalami, Karen
  Mar-blestone, Sarah Schuster, and Karen Stevens.
\newblock {Why Do Alzheimer Patients Have Difficulty with Pronouns? Working
  Memory, Semantics, and Reference in Comprehension and Production in
  Alzheimer's Disease}.
\newblock {\em Brain and Language}, 67:202--227, 1999.

\bibitem{Wendelstein2015}
Britta Wendelstein, J{\"{o}}rn Stegmeier, Claudia Frankenberg, Ekkehard Felder,
  and Johannes Schr{\"{o}}der.
\newblock {Changes in the use of pronouns in spoken language in the course of
  preclinical to early stages of Alzheimer's disease}.
\newblock {\em Alzheimer's {\&} Dementia}, 11(7):P784--P785, jul 2015.

\bibitem{Jarrold2014}
William Jarrold, Bart Peintner, David Wilkins, Dimitra Vergryi, Colleen Richey,
  Maria~Luisa Gorno-Tempini, and Jennifer Ogar.
\newblock {Aided diagnosis of dementia type through computer-based analysis of
  spontaneous speech}.
\newblock {\em Proceedings of the Workshop on Computational Linguistics and
  Clinical Psychology: From Linguistic Signal to Clinical Reality}, pages
  27--37, 2014.

\bibitem{Wendelstein2014}
Britta Wendelstein, Ekkehard Felder, J{\"{o}}rn Stegmeier, and Johannes
  Schr{\"{o}}der.
\newblock {Noun or Pronoun?: Using Parts of Speech in ALzheimer's Disease}.
\newblock {\em Alzheimer's {\&} Dementia}, 10(4):P434, jul 2014.

\end{thebibliography}
